# Positional number systems

```{r}
#| label: setup

library(tibble)
library(gt)
```

::: {#def-positional-notation}

## Positional notation

Positional notation using base $b$ (or radix $b$) is defined by the rule [@knuth_art_1997, p. 195]:

$$(a_na_{n-1} \ldots a_1a_0.a_{-1}a_{-2}a_{-3} \ldots)_b= \sum_{k=0}^n a_kb^k + \sum_{k = 1}^{\infty} a_{-k}b^{-k}$$
Where the $a$'s are called the digits of the representation.

Also $0 \leq a_k < b$, $0 \leq a_{-k} < b$ and $b^k, b^{-k}$ are the weights, $w$, of the corresponding digits where the position $k$ is the base $b$ logarithm of the corresponding $w$, that is $k = \log_b w = \log_b b^k$. 

Furthermore, the dot, $.$ , that appears between $a_0$ and $a_{-1}$ is called the radix point were it is used to identify the integer part from the fractional part of the representation. In that sense, $a_n$ to $a_0$ is the integer part follow by the radix point and then $a_{-1}, a_{-2}, a_{-3}, \ldots$ is the fractional part.

:::

::: {#exm-positional-representation}

## For $(5627.35)_{10}$

We have that $b=10$ and $(5627.35)_{10} = 5 \cdot 10^3 + 6 \cdot 10^2 + 2 \cdot 10^1 + 7 \cdot 10^0 + 3 \cdot 10^{-1} + 5 \cdot 10^{-2} + 0 \cdot 10^{-3} + \cdots$ where:

```{r}
#| label: tbl-positional-notation-example
#| tbl-cap: Positional notation example

pos_not_ex <- data.frame(
 desc    = c("Position",
             "Weight",
             "Digit",
             "Example weight",
             "Example digit"),
 pos_3   = c("$3$", "$b^3$", "$a_3$", "$10^3$", "5"),
 pos_2   = c("$2$", "$b^2$", "$a_2$", "$10^2$", "6"),
 pos_1   = c("$1$", "$b^1$", "$a_1$", "$10^1$", "2"),
 pos_0   = c("$0$", "$b^0$", "$a_0$", "$10^0$", "7"),
 pos_d_1 = c("$-1$", "$b^{-1}$", "$a_{-1}$", "$10^{-1}$", "3"),
 pos_d_2 = c("$-2$", "$b^{-2}$", "$a_{-2}$", "$10^{-2}$", "5"),
 pos_d_3 = c("$-3$", "$b^{-3}$", "$a_{-3}$", "$10^{-3}$", "0"),
 pos_d_  = rep("$\\cdots$", times = 5))

gt(pos_not_ex) |> 
 tab_options(column_labels.hidden = TRUE) |> 
 fmt_markdown()
```

:::

## Base conversion [@britton_base_2017]

### Convert from base $10$ to base $b$

#### Integer part

1. Divide the integer part by $b$ and record the remainder
2. Take the quotient from the previous division and divide it by $b$, recording the new remainder
3. Repeat step 2. until the quotient becomes $0$
4. The integer part in base $b$ is the remainders written in reverse order

#### Fraction part

1. Multiply the fraction part by $b$ and record the integer part
2. Take the fractional part of the result and multiply it by $b$, then record the new integer part
3. Repeat step 2. until the fractional part becomes $0$ or you have reached the desired number of digits to avoid infinite loops in repeating decimals
4. The fractional part in base $b$ is the sequence of recorded integers written in order

::: {#exm-base-10-to-base-2}

## For $(45.3689)_{10}$ to base $2$ with 5 decimal places

In base $2$ the digits can be defined as $\{ 1, 2 \}$.

```{r}
#| label: tbl-exm-base-10-to-base-2-int
#| tbl-cap: For $(45)_{10}$  

from_base_10_to_base_2_int <- tibble(
 Step = 1:6,
 Division = c(
  "$\\frac{45}{2}$",
  "$\\frac{22}{2}$",
  "$\\frac{11}{2}$",
  "$\\frac{5}{2}$",
  "$\\frac{2}{2}$",
  "$\\frac{1}{2}$"
 ),
 Quotient = c(22, 11, 5, 2, 1, 0),
 Remainder = c(1, 0, 1, 1, 0, 1)
)

from_base_10_to_base_2_int |> 
 gt() |> 
 tab_header(title = "Integer part") |> 
 fmt_markdown()
```

```{r}
#| label: tbl-exm-base-10-to-base-2-frac
#| tbl-cap: For $(0.3689)_{10}$

from_base_10_to_base_2_frac <- tibble(
 Step = 1:5,
 Multiplication = c(
  "$0.3689 \\times 2 = 0.7378$",
  "$0.7378 \\times 2 = 1.4756$",
  "$0.4756 \\times 2 = 0.9512$",
  "$0.9512 \\times 2 = 1.9024$",
  "$0.9024 \\times 2 = 1.8048$"
 ),
 Integer = c(0, 1, 0, 1, 1),
 Fraction = c(0.7378, 0.4756, 0.9512, 0.9024, 0.8048)
)

from_base_10_to_base_2_frac |> 
 gt() |> 
 tab_header(title = "Fractional part") |> 
 fmt_markdown()
```

Therefore $(45.3689)_{10} \approx (101101.01011)_2$ 

:::

::: {#exm-base-10-to-base-16}

## For $(1251.9324)_{10}$ to base $16$ with 6 decimal places

In base $16$ the digits can be defined as $\{ 1, 2, \ldots, 8, 9, A, B, C, D, E, F \}$.

```{r}
#| label: tbl-exm-base-10-to-base-16-int
#| tbl-cap: For $(1251)_{10}$ 

from_base_10_to_base_16_int <- tibble(
 Step = 1:3,
 Division = c(
  "$\\frac{1251}{16}$",
  "$\\frac{78}{16}$",
  "$\\frac{4}{16}$"
 ),
 Quotient = c(78, 4, 0),
 Remainder = c(3, 14, 4),
 `16 base digit` = c("3", "E", "4")
)

from_base_10_to_base_16_int |> 
 gt() |> 
 tab_header(title = "Integer part") |> 
 fmt_markdown()
```

```{r}
#| label: tbl-exm-base-10-to-base-16-frac
#| tbl-cap: For $(0.9324)_{10}$

from_base_10_to_base_16_frac <- tibble(
 Step = 1:6,
 Multiplication = c(
  "$0.9324 \\times 16 = 14.9184$",
  "$0.9184 \\times 16 = 14.6944$",
  "$0.6944 \\times 16 = 11.1104$",
  "$0.1104 \\times 16 = 1.7664$",
  "$0.7664 \\times 16 = 12.2624$",
  "$0.2624 \\times 16 = 4.1984$"
 ),
 Integer = c(14, 14, 11, 1, 12, 4),
 `16 base digit` = c("E", "E", "B", "1", "C", "4"),
 Fraction = c(0.9184, 0.6944, 0.1104, 0.7664, 0.2624, 0.1984)
)

from_base_10_to_base_16_frac |> 
 gt() |> 
 tab_header(title = "Integer part") |> 
 fmt_markdown()
```

Therefore $(1251.9324)_{10} \approx (4E3.EEB1C4)_{16}$

:::

### Convert from base $b$ to base $10$

#### Integer part

\begin{align*}
 (a_na_{n-1} \ldots a_1a_0)_b & =  a_n \cdot b^n + a_{n-1} \cdot b^{n-1} + \dots  a_1 \cdot b^1 + a_0 \cdot b^0 \\
 & = \left( \sum_{k=0}^n a_kb^k \right)_{10}
\end{align*}

#### Fraction part

\begin{align*}
 (a_{-1}a_{-2}a_{-3} \cdots)_b & =  a_{-1} \cdot b^{-1} + a_{-2} \cdot b^{-2} + a_{-3} \cdot b^{-3} + \ldots \\
 & = \left( \sum_{k=1}^{\infty} a_{-k}b^{-k} \right)_{10}
\end{align*}

::: {#exm-base-3-to-base-10}

## For $(12011.1022)_3$ to base $10$ with 4 decimal places

In base $3$ the digits can be defined as $\{ 1, 2, 3 \}$.

- Integer part

\begin{align*}
 (12011)_3 & =  1 \cdot 3^4 + 2 \cdot 3^3 + 0 \cdot 3^2 + 1 \cdot 3^1 + 1 \cdot 3^0 \\
 & = 81 + 54 + 0 + 3 + 1 \\
 & = (139)_{10}
\end{align*}

- Fraction part

\begin{align*}
 (0.1022)_3 & =  1 \cdot 3^{-1} + 0 \cdot 3^{-2} + 2 \cdot 3^{-3} +  2 \cdot 3^{-4} \\
 & = \frac{1}{3} + 0 + \frac{2}{27} + \frac{2}{81} \\
 & \approx 0.3333 + 0 + 0.0741 + 0.0247 \\
 & = 0.4321
\end{align*}

Therefore $(12011.1022)_3 \approx (139.4321)_{10}$

:::

::: {#exm-base-14-to-base-10}

## For $(C4B3.3A9D)_14$ to base $10$ with 5 decimal places

In base $14$ the digits can be defined as $\{ 1, 2, \ldots, 8, 9, A, B, C, D \}$.

- Integer part

\begin{align*}
 (C4B3)_{14} & =  C \cdot 14^3 + 4 \cdot 14^2 + B \cdot 14^1 + 3 \cdot 14^0 \\
 & = 12 \cdot 14^3 + 4 \cdot 14^2 + 11 \cdot 14^1 + 3 \cdot 14^0 \\
 & = 12 \cdot 2744 + 4 \cdot 196 + 11 \cdot 14 + 3 \\
 & = (33869)_{10}
\end{align*}

- Fraction part

\begin{align*}
 (0.3A9D)_{14} & =  3 \cdot 14^{-1} + A \cdot 14^{-2} + 9 \cdot 14^{-3} +  D \cdot 14^{-4} \\
 & = \frac{3}{14} + \frac{10}{196} + \frac{9}{2744} + \frac{13}{38416} \\
 & \approx 0.21429 + 0.05102 + 0.00328 + 0.00034 \\
 & = 0.26893
\end{align*}

Therefore $(12011.1022)_{14} \approx (33869.26893)_{10}$

:::

### Convert from base $b$ to base $a$

1. Convert from base $b$ to base $10$
2. Convert from base $10$ to base $a$

