# Positional number systems

```{r}
#| label: setup

library(gt)
```

::: {#def-positional-notation}

## Positional notation

Positional notation using base $b$ (or radix $b$) is defined by the rule:

$$(a_na_{n-1} \ldots a_1a_0.a_{-1}a_{-2}a_{-3})_b = \sum_{k=0}^n a_kb^k + \sum_{k = 1}^{\infty} a_{-k}b^{-k}$$
Where the $a$'s are called the digits of the representation.

Also $0 \leq a_k < b$, $0 \leq a_{-k} < b$ and $b^k, b^{-k}$ are the weights, $w$, of the corresponding digits where the position $k$ is the base $b$ logarithm of the corresponding $w$, that is $k = \log_b w = \log_b b^k$. 

Furthermore, the dot, $.$,that appears between $a_0$ and $a_{-1}$ is called the radix point were it is used to identify the integer part from the fractional part of the representation. In that sense, $a_n$ to $a_0$ is the integer part follow by the radix point and then $a_{-1}, a_{-2}, a_{-3}, \ldots$ is the fractional part.

:::

For example for $(5627.35)_{10}$ we have that $b=10$ and $(4327.33)_{10} = 5 \cdot 10^3 + 6 \cdot 10^2 + 2 \cdot 10^1 + 7 \cdot 10^0 + 3 \cdot 10^{-1} + 5 \cdot 10^{-2} + 0 \cdot 10^{-3} + \cdots$ where:

```{r}
#| label: tbl-positional-notation-example
#| tbl-cap: Positional notation example

pos_not_ex <- data.frame(desc    = c("Position",
                                     "Weight",
                                     "Digit",
                                     "Example weight",
                                     "Example digit"),
                         pos_3   = c("$3$", "$b^3$", "$a_3$", "$10^3$", "5"),
                         pos_2   = c("$2$", "$b^2$", "$a_2$", "$10^2$", "6"),
                         pos_1   = c("$1$", "$b^1$", "$a_1$", "$10^1$", "2"),
                         pos_0   = c("$0$", "$b^0$", "$a_0$", "$10^0$", "7"),
                         pos_d_1 = c("$-1$", "$b^{-1}$", "$a_{-1}$", "$10^{-1}$", "3"),
                         pos_d_2 = c("$-2$", "$b^{-2}$", "$a_{-2}$", "$10^{-2}$", "5"),
                         pos_d_3 = c("$-3$", "$b^{-3}$", "$a_{-3}$", "$10^{-3}$", "0"),
                         pos_d_  = rep("$\\cdots$", times = 5))

gt(pos_not_ex) |> 
 tab_options(column_labels.hidden = TRUE) |> 
 fmt_markdown()
```
